---
title: "R Basics"
author: "Kai Nguyen"
date: "2025-06-06"
output:
  html_document: default
---
# Conventions
```{r}
# This is an edited transcript of a R session
# We've inserted comment lines, which begin with a '#'
```
_Command lines_ begin with a '>'. You should enter everything after the '>' and hit return. R's response will be in the line or lines below the command.  

***Trick***: You can use the up arrow to find previous commands.  
  
# Using a tutorial  
  
  In order to (begin to) learn the commands you should have RStudio open and go through the tutorial line-by-line. When the tutorial shows you a command try it in RStudio. Then make up a few of your own variants and try them. If you wonder about the effect of some command, try it. The worst that will happen is R will print out an error message.  
  
# Introduction  
  
  R is a full featured statistics package as well as a full programming language. The command window is the one with the >. It's probably the window in the lower left.  
  
# Starting R  
  
  If you've installed R and RStudio then simply start the RStudio application. The command window is the one with the >. It's probably the window in the lower left.  
  
# R as a Calculator  
  
  <span style="color:red;"># The basic operations are *, +, -, /, ^</span>  
  


```{r}
2+3
2*3
2/3
2^3
2*(3+1)^2
```
  
  Don't worry for now about the [1]. It will play more of a role when R is printing numbers in a list.
  
# Using Variables  
  
  <span style="color:red;"># You can store results in variables and use them in calculations.</span>  
  
```{r}
x = 2 + 3
```
<span style="color:red;"># When you assign a value to a variable it does not echo the answer to screen. You can see the value of x by just using x as a command.</span>  
```{r}
print(x)
y = 1 + 2
print(x*y)
z = x^y
print(z)
```
<span style="color:red;"># R has another notation for assignment: the arrow: <-. Many R programmers use this. It may seem odd to programmers coming from other languages.</span> 

```{r}
x <- 3
print(x)
x <- 5.412
print(x)
```

# Vectors  
### A vector is a type of list. Often it is a list of numbers, but it can be a list of other types such as characters.
### You creat vectors by using the c().  
  
   <span style="color:red;"># A vector with 4 entries</span>  
   
   <span style="color:red;"># (Of course, you can store vectors in variables.)</span>
```{r}
x = c(1.1, 0.0, 3.14, 2.718)
print(x)
```

<span style="color:red;"># Of course using the arrow instead of equal sign works here.</span>
```{r}
x <- c(2,4,6)
print(x)
```

<span style="color:red;"># Sequences of intergers are so common that there is a shortcut for making them.</span>
```{r}
x = 1:4
print(x)
x = 3:10
print(x)
x = 9:2
print(x)
```

<span style="color:red;"># A long vector will be displayed over several lines. At the start of each line in brackets is the index of the first entry on that line.</span>
```{r}
x = 1:40
print(x)
```

# Vector Arithmetic  
  
### You can add a number to a vector. This means adding the number to every element of the vector.
```{r}
x = c(1,3,5)
y = x + 7.1
print(y)
```

### Subtraction, multiplication, division and powers work the same way.
```{r}
x= c(1,3,5)
#Multiplication
y = 7*x
print(y)

#Division
y =x/7
print(y)

#Powers
y=x^6
print(y)
```

### You can add, subtract, multiply, and divide vectors of the same size.
```{r}
# Given 2 vectors x and y
x = c(1, 2, 3)
y = c(4, 5, 6)

#add
print(x+y)

#subtract
print(x-y)

#multiply
print(x*y)

#divide
print(x/y)

```
### You can even raise a vector to another vector of the same length
```{r}
#raise a vector x^y
print(x^y)
```

# Accessing entries in a vector.  
### Entries in vectors are found with the notation x[j]
```{r}
x = c(2, 4, 6, 8, 10)
print(x[1])
print(x[2])
print(x[3])
print(x[4])
```
### R allows you to access more than one entry at a time
 x has 8 elements
```{r}
x = 2*c(1, 2, 3, 4, 5, 6, 7, 8)
print(x)
```

### Notice that we have to put a vector of indices inside the brackets to access the first and second entries in x
```{r}
y = x[c(1,2)]
print(y)
y = x[c(1,3,6)]
print(y)
```

### We can access the same entry multiple times.
```{r}
y = x[c(2,2,2,1)]
print(y)
```

### Using the colon method of creating vectors is useful here.
```{r}
y = x[2:5]
print(y)
```

# Functions on Vectors
***R has all the functions you know and love. Most of them can be used on vectors.***

### We will start with functions numbers
```{r}
x = sin(1)
print(x)

x = sin(2)
print(x)

x = cos(112/4)
print(x)
```

### R knows about pi
```{r}
print(pi)
print(tan(pi/4))
```

### The exponential function is given by 'exp'
```{r}
x = exp(0)
print(x)
x = exp(1)
print(x)
```

### Sin acts on vectors by taking the sin of each element.
```{r}
x = c(1, 2, 3, pi/2)
print(x)
y = sin(x)
print(y)
```


### exp also acts on vectors
```{r}
x = c(0, 2, 1)
y = exp(x)
print(y)

y = exp(x[c(1, 2)])
print(y)
```
### The sum() and mean() functions on vectors take the sum and average respectively of the vectors entries.

```{r}
# given vector x(1,2,3,4,5,6)
x = 1:6
print(x)

# calculate sum of all entries of vector x
s = sum(x)
print(s)

# calculate average of all x's entries
m = mean(x)
print(m)

```

### Example: find the sum of the integers from 1 to 1024.
```{r}
x = 1:1024
s = sum(x)
print(s)
```

### This can be done in one command
```{r}
s = sum(1:1024)
print(s)
```

# A few more examples with powers
### Example: find the sum of the squares of the integers from 1 to 1024 
```{r}
x = 1:1024
s = sum(x^2)
print(s)

# In one line
print(sum((1:1024)^2))
```

# Matrices
## R's method of creating matrices is a little bit clunky. It works by arranging the entries in a vector into rows and columns.  
  
### A vector of length 10 can be arranged as a 2x5 or a 5x2 matrix
### Again the syntax is clunky but very clear
```{r}
#Vector x(1,2,3,4,5,6,7,8,9,10)
x = 1:10
print(x)

#Matrix y 2x5(2 row 5 column)
y = matrix(x, nrow=2, ncol=5)
print(y)

#Matrix z 5x2(5 row 2 column)
z = matrix(x, nrow=5, ncol=2)
print(z)

```

<span style="color:red;"># Notice in the examples above that R builds the matrices one column at a time. That is, our vector 1 to 10 runs in sequence down the column.</span>  
  
<span style="color:red;"># However the matrix() function, like most R function, has a lot of optional parameters that allow you to control its behavior.</span>

### How to make it build a matrix by rows
```{r}
y = matrix(x, nrow = 2, ncol = 5, byrow = TRUE)
print(y)
```
# Accessing Entries in Matrices
### The square bracket notation is used by giving both the row and column indices.
```{r}
x = 1:10
y = matrix(x, nrow=2, ncol=5)
print(y)

#print an element or an entry
print(y[1,2])
print(y[2,3])

```

# Sums and Means on Matrices
### Taking the sums or means of the columns(or rows) of a matrix will be every useful to us
```{r}
# Start by creating a matrix to practice on.
x = 1:10
y = matrix(x, nrow=2, ncol=5)
print(y)

#To sum each of the columns we use the colSums() function.
# y has five columns so colSums(y) produces 5 numbers.
s = colSums(y)
print(s)

# y has two rows so rowSums(y) produces 2 numbers
s = rowSums(y)
print(s)

# Likewise rowMeans(y) and colMeans(y)
m = rowMeans(y)
print(m)
m = colMeans(y)
print(m)

```

